### 2020-04-14 2020.csic5011/homework/homework06.tex
homework06

\lecture{Homework 4. ISOMAP \& LLE}{Yuan Yao}{Due: Tuesday October 28, 2014}{October 21, 2014}

@@

@@

To overcome this hurdle, Nystr\"{o}m method leads us to a scalable approach to compute eigenvectors of low rank matrices.  
Suppose that an $N$-by-$N$ positive semidefinite matrix $K\succeq 0$ admits the following block partition
\begin{equation}
K = \left[
\begin{array}{cc}
A & B \\
B^T & C
\end{array}
\right].
\end{equation}
where $A$ is an $n$-by-$n$ block. Assume that $A$ has the spectral decomposition $A=U \Lambda U^T$, $\Lambda=\diag(\lambda_i)$ ($\lambda_1\geq \lambda_2 \geq \ldots \lambda_k > \lambda_{k+1}=\ldots=0$) and $U=[u_1,\ldots,u_n]$ satisfies $U^TU=I$. 
\begin{enumerate}
\item Assume that $K=X X^T$ for some $X=[X_1; X_2] \in \R^{N\times k}$ with the block $X_1\in \R^{n\times k}$. Show that $X_1$ and $X_2$ can be decided by:
\begin{equation}\label{eq:x1}
 X_1 = U_k \Lambda_k^{1/2}, 
\end{equation}
\begin{equation} \label{eq:x2}
X_2 = B^T U_k \Lambda_k^{-1/2}, 
\end{equation}
%\[ X_1 X_1^T = A \Rightarrow X_1 = U_k \Lambda_k^{1/2}, \]
%\[ X_1 X_2^T = B \Rightarrow X_2 = B^T U_k \Lambda_k^{-1/2}. \]  
where $U_k=[u_1, \ldots, u_k]$ consists of those $k$ columns of $U$ corresponding to top $k$ eigenvalues $\lambda_i$ ($i=1,\ldots,k$).  
\item Show that for general $K\succeq 0$, one can construct an approximation from \eqref{eq:x1} and \eqref{eq:x2}, 
\begin{equation}
\hat{K} = \left[
\begin{array}{cc}
A & B \\
B^T & \hat{C} 
\end{array}
\right].
\end{equation}
where $A=X_1 X_1^T$, $B= X_1 X_2^T$, and $\hat{C}=X_2 X_2^T =   B^T A^\dagger B$, $A^\dagger$ denoting the Moore-Penrose (pseudo-) inverse of $A$. Therefore $\|\hat{K}-K\|_F =\|C- B^T A^\dagger B\|_F$. Here the matrix $C-B^T A^\dagger B=:K/A$ is called the (generalized) \emph{Schur Complement} of $A$ in $K$.
\item Explore Nystr\"{o}m method on the Swiss-Roll dataset (\url{http://yao-lab.github.io/data/swiss_roll_data.mat} contains 3D-data X; \url{http://yao-lab.github.io/data/swissroll.m} is the matlab code) with ISOMAP. To construct the block $A$, you may choose either of the following: 
\subitem $n$ random data points; 
\subitem *$n$ landmarks as minimax $k$-centers (\url{https://yao-lab.github.io/data/kcenter.m}); \\
Some references can be found at: 
\subitem[dVT04] Vin de Silva and J. B. Tenenbaum, ``Sparse multidimensional scaling using landmark points", 2004, downloadable at \url{http://pages.pomona.edu/~vds04747/public/papers/landmarks.pdf};  
\subitem[P05] John C. Platt, ``FastMap, MetricMap, and Landmark MDS are all Nystr\"{o}m Algorithms", 2005, downloadable at \url{http://research.microsoft.com/en-us/um/people/jplatt/nystrom2.pdf}. 
\item *Assume that $A$ is invertible, show that  
\[ \det(K) = \det(A)\cdot \det(K/A), \]
\item *Assume that $A$ is invertible, show that 
\[ \rank(K)=\rank(A)+\rank(K/A). \]
\item *Can you extend the identities in (c) and (d) to the case of noninvertible $A$? A good reference can be found at, 
\subitem[Q81] Diane V. Quellette, ``Schur Complements and Statistics", Linear Algebra and Its Applications, 36:187-295, 1981. \url{http://www.sciencedirect.com/science/article/pii/0024379581902329}
\end{enumerate}

>>>>

\lecture{Homework 6. Manifold Learning}{Yuan Yao}{Due: 1.5 weeks later}{Apr. 8, 2020}

@@

\item {\em Manifold Learning}: The following codes by Todd Wittman contain major manifold learning algorithms talked on class.

%\url{http://www.math.pku.edu.cn/teachers/yaoy/Spring2011/matlab/mani.m}
\url{http://math.stanford.edu/~yuany/course/data/mani.m}

Precisely, eight algorithms are implemented in the codes: MDS, PCA, ISOMAP, LLE, Hessian Eigenmap, Laplacian Eigenmap, Diffusion Map, and LTSA. 
The following nine examples are given to compare these methods,
\begin{enumerate}
\item Swiss roll;
\item Swiss hole;
\item Corner Planes;
\item Punctured Sphere;
\item Twin Peaks;
\item 3D Clusters;
\item Toroidal Helix;
\item Gaussian;
\item Occluded Disks.
\end{enumerate}
Run the codes for each of the nine examples, and analyze the phenomena you observed.

\subitem *Moreover if possible, play with t-SNE using scikit-learn manifold package:

\url{http://scikit-learn.org/stable/modules/generated/sklearn.manifold.TSNE.html}

or any other implementations collected at

\centering{\url{http://lvdmaaten.github.io/tsne/} }

%\subitem* Moreover if possible, make an implementation of Vector Diffusion Map (Laplacian) to reconstruct Swiss roll etc., and compare it against others. Note that LTSA looks for the following coordinates,
%\[ \min_{Y}  \sum_{i\sim j} \| y_i - U_i U_j^T y_j \|^2 \]
%where $U_i$ is a local PCA basis for tangent space at point $x_i\in \R^p$; in a contrast, vector connection Laplacian looks for:
%\[ \min_{Y}  \sum_{i\sim j} \| y_i - O_{ij} y_j \|^2, \ \ \ O_{ij} = \arg \min_O \|U_i - O_{ij} U_j \|^2 \]
%where $O_{ij}$ is an isometry (parallel transportation) from the tangent space at point $x_j$ to the tangent space at point $x_i$, whose solution is given by the lower eigenvectors of the following vector connection Laplacian
%\[
%L_{ij} = \left\{
%\begin{array}{lr}
%-w_{ij} O_{ij}, & i\neq j \\
%\sum_{j} w_{ij} I, & i=j.
%\end{array}
%\right.
%\]
%Here $w_{ij}=w_{ji}$ defines the similarity (neighborhood) between point $x_i$ and $x_j$. For details, please refer to the paper by Amit Singer and Hau-Tieng Wu, \emph{Vector diffusion maps and the connection laplacian}, Comm. Pure Appl. Math. 65 (2012), no. 8, 1067–1144. You may try Hau-Tieng Wu's matlab codes on VDM from the following website
%
%\centering{\url{https://sites.google.com/site/hautiengwu/home/download}}

@@

To overcome this hurdle, Nystr\"{o}m method leads us to a scalable approach to compute eigenvectors of low rank matrices.  
Suppose that an $N$-by-$N$ positive semidefinite matrix $K\succeq 0$ admits the following block partition
\begin{equation}
K = \left[
\begin{array}{cc}
A & B \\
B^T & C
\end{array}
\right].
\end{equation}
where $A$ is an $n$-by-$n$ block. Assume that $A$ has the spectral decomposition $A=U \Lambda U^T$, $\Lambda=\diag(\lambda_i)$ ($\lambda_1\geq \lambda_2 \geq \ldots \lambda_k > \lambda_{k+1}=\ldots=0$) and $U=[u_1,\ldots,u_n]$ satisfies $U^TU=I$. 
\begin{enumerate}
\item Assume that $K=X X^T$ for some $X=[X_1; X_2] \in \R^{N\times k}$ with the block $X_1\in \R^{n\times k}$. Show that $X_1$ and $X_2$ can be decided by:
\begin{equation}\label{eq:x1}
 X_1 = U_k \Lambda_k^{1/2}, 
\end{equation}
\begin{equation} \label{eq:x2}
X_2 = B^T U_k \Lambda_k^{-1/2}, 
\end{equation}
%\[ X_1 X_1^T = A \Rightarrow X_1 = U_k \Lambda_k^{1/2}, \]
%\[ X_1 X_2^T = B \Rightarrow X_2 = B^T U_k \Lambda_k^{-1/2}. \]  
where $U_k=[u_1, \ldots, u_k]$ consists of those $k$ columns of $U$ corresponding to top $k$ eigenvalues $\lambda_i$ ($i=1,\ldots,k$).  
\item Show that for general $K\succeq 0$, one can construct an approximation from \eqref{eq:x1} and \eqref{eq:x2}, 
\begin{equation}
\hat{K} = \left[
\begin{array}{cc}
A & B \\
B^T & \hat{C} 
\end{array}
\right].
\end{equation}
where $A=X_1 X_1^T$, $B= X_1 X_2^T$, and $\hat{C}=X_2 X_2^T =   B^T A^\dagger B$, $A^\dagger$ denoting the Moore-Penrose (pseudo-) inverse of $A$. Therefore $\|\hat{K}-K\|_F =\|C- B^T A^\dagger B\|_F$. Here the matrix $C-B^T A^\dagger B=:K/A$ is called the (generalized) \emph{Schur Complement} of $A$ in $K$.
\item Explore Nystr\"{o}m method on the Swiss-Roll dataset (\url{http://yao-lab.github.io/data/swiss_roll_data.mat} contains 3D-data X; \url{http://yao-lab.github.io/data/swissroll.m} is the matlab code) with ISOMAP. To construct the block $A$, you may choose either of the following: 
\subitem $n$ random data points; 
\subitem *$n$ landmarks as minimax $k$-centers (\url{https://yao-lab.github.io/data/kcenter.m}); \\
Some references can be found at: 
\subitem[dVT04] Vin de Silva and J. B. Tenenbaum, ``Sparse multidimensional scaling using landmark points", 2004, downloadable at \url{http://pages.pomona.edu/~vds04747/public/papers/landmarks.pdf};  
\subitem[P05] John C. Platt, ``FastMap, MetricMap, and Landmark MDS are all Nystr\"{o}m Algorithms", 2005, downloadable at \url{http://research.microsoft.com/en-us/um/people/jplatt/nystrom2.pdf}. 
\item *Assume that $A$ is invertible, show that  
\[ \det(K) = \det(A)\cdot \det(K/A), \]
\item *Assume that $A$ is invertible, show that 
\[ \rank(K)=\rank(A)+\rank(K/A). \]
\item *Can you extend the identities in (c) and (d) to the case of noninvertible $A$? A good reference can be found at, 
\subitem[Q81] Diane V. Quellette, ``Schur Complements and Statistics", Linear Algebra and Its Applications, 36:187-295, 1981. \url{http://www.sciencedirect.com/science/article/pii/0024379581902329}

%$$
%\begin{array}{ll}{{\text { minimize }}} & {\|L\|_{*}+\lambda\|S\|_{1}} \\ {\text { subject to }} & {L+S=M}\end{array}
%$$
